---
title: "repdata2"
author: "Kent Mok"
date: "August 19, 2015"
output: html_document
---

# Weather Events in the United States

## Synopsis
This analysis reviews storm data provided by the U. S. National Oceanic and 
Atmospheric Administration's (NOAA) storm database. The database tracks 
characteristics of major storms and weather events in the United States, 
including when and where they occur, as well as estimates of any fatalities, 
injuries, and property damage. Using the data, this analysis determines which
type of events are the most harmful to people by looking at fatalities and 
injuries related to the events. In addition, the analysis determines which type
of events have the greatest economic consequences.

## Data Processing
```{r}
# Load data
rawdata <- read.csv("repdata-data-StormData.csv")
```

The data is quite messy as can be seen by looking at unique values of EVTYPE in
the data. The data needs to be cleaned to match the standard Storm Data Event
values set forth by the NWS.
```{r}
levels(rawdata$EVTYPE)
```

The below code is used to clean the data based on event types.
```{r}
# Clean data
rawdata$EVTYPE[grep("HIGH SURF", rawdata$EVTYPE)] <- "HIGH SURF"
rawdata$EVTYPE[grep("HEAVY SURF", rawdata$EVTYPE, ignore.case = TRUE)] <- 
    "HEAVY SURF"
rawdata$EVTYPE[grep("RIP CURRENT", rawdata$EVTYPE)] <- "RIP CURRENT"
# "Burst" included under Thunderstorm Wind per 7.39.1 of NWSI 10-1605
# According to the remarks column, all EVTYPE's beginning with SUMMARY were
# related to thunderstorms.
tstormmatch <- c("THUNDERSTORM WIND", "THUNDERSTORMS WIND",
                 "THUDERSTORM WINDS", "THUNDEERSTORM WINDS",
                 "THUNDERESTORM WINDS", "THUNDERSTORM  WINDS",
                 "THUNDERSTORM W INDS", "THUNDERSTORM WINS",
                 "THUNDERSTORMW WINDS", "THUNDERSTORMWINDS",
                 "THUNDERSTROM WIND", "THUNDERSTROM WINDS",
                 "THUNDERTORM WINDS", "THUNDERTSORM WIND",
                 "THUNDESTORM WINDS", "THUNERSTORM WINDS", "BURST",
                 "TUNDERSTORM WIND", "^Summary", "^TSTM", "^ TSTM")
rawdata$EVTYPE[grep(paste(tstormmatch, collapse = "|"), rawdata$EVTYPE,
    ignore.case = TRUE)] <- "THUNDERSTORM WIND"
rawdata$EVTYPE[grep("^FLASH", rawdata$EVTYPE)] <- "FLASH FLOOD"
rawdata$EVTYPE[grep("^FLOOD", rawdata$EVTYPE)] <- "FLOOD"
rawdata$EVTYPE[grep("WILD", rawdata$EVTYPE)] <- "WILDFIRE"
lightningmatch <- c("^LIGHTNING", "^ LIGHTNING", "RED FLAG CRITERIA")
rawdata$EVTYPE[grep(paste(lightningmatch, collapse = "|"), rawdata$EVTYPE)] <-
    "LIGHTNING"
rawdata$EVTYPE[grep("^HAIL", rawdata$EVTYPE)] <- "HAIL"
dustdevilmatch <- c("DUST DEVIL", "DUST DEVEL")
rawdata$EVTYPE[grep(paste(dustdevilmatch, collapse = "|"), rawdata$EVTYPE,
    ignore.case = TRUE)] <- "DUST DEVIL"
rawdata$EVTYPE[grep("VOLCANIC", rawdata$EVTYPE, ignore.case = TRUE)] <- 
    "VOLCANIC ASH"
rawdata$EVTYPE[grep("^HEAVY SNOW", rawdata$EVTYPE, ignore.case = TRUE)] <- 
    "HEAVY SNOW"
rawdata$EVTYPE[grep("HURRICANE", rawdata$EVTYPE, ignore.case = TRUE)] <- 
    "HURRICANE/TYPHOON"
waterspoutmatch <- c("^WATERSPOUT", "^ WATERSPOUT", "WAYTERSPOUT",
                     "WATER SPOUT", "LANDSPOUT")
rawdata$EVTYPE[grep(paste(waterspoutmatch, collapse = "|"), rawdata$EVTYPE,
    ignore.case = TRUE)] <- "WATERSPOUT"
densefogmatch <- c("DENSEFOG", "^FOG")
rawdata$EVTYPE[grep(paste(densefogmatch, collapse = "|"), rawdata$EVTYPE,
    ignore.case = TRUE)] <- "DENSE FOG"
freezefogmatch <- c("FREEZING FOG", "ICE FOG")
rawdata$EVTYPE[grep(paste(freezefogmatch, collapse = "|"), rawdata$EVTYPE,
    ignore.case = TRUE)] <- "FREEZING FOG"
heavyrainmatch <- c("^HEAVY RAIN", "LOCALLY HEAVY RAIN", "TSTM HEAVY RAIN")
rawdata$EVTYPE[grep(paste(heavyrainmatch, collapse = "|"), rawdata$EVTYPE,
    ignore.case = TRUE)] <- "HEAVY RAIN"
# "Freezing drizzle" included under Winter Weather per 7.48 of NWSI 10-1605
# Also assuming that "Wintry" falls under this category
wintweatmatch <- c("WINTER WEATHER", "FREEZING DRIZZLE", "WINTERY", "WINTRY",
                   "WINTER MIX")
rawdata$EVTYPE[grep(paste(wintweatmatch, collapse = "|"), rawdata$EVTYPE,
    ignore.case = TRUE)] <- "WINTER WEATHER"
rawdata$EVTYPE[grep("TROPICAL STORM", rawdata$EVTYPE, ignore.case = TRUE)] <- 
    "TROPICAL STORM"
# "Freezing rain" included under Winter Storm per 7.47 of NWSI 10-1605
# "Light snow" included under Winter Storm per 7.47 of NWSI 10-1605
wstormmatch <- c("^WINTER STORM", "^FREEZING RAIN", "LIGHT SNOW")
rawdata$EVTYPE[grep(paste(wstormmatch, collapse = "|"), rawdata$EVTYPE, 
    ignore.case = TRUE)] <- "WINTER STORM"
rawdata$EVTYPE[grep("^STRONG WIND", rawdata$EVTYPE, ignore.case = TRUE)] <- 
    "STRONG WIND"
# Since "High Winds" accompany blizzards, any event type with high wind and
# blizzard in it will be resolved to just "Blizzard"
rawdata$EVTYPE[grep("BLIZZARD", rawdata$EVTYPE, ignore.case = TRUE)] <- 
    "BLIZZARD"
# Any event type with "Drought" resolved to just "Drought"
rawdata$EVTYPE[grep("DROUGHT", rawdata$EVTYPE, ignore.case = TRUE)] <- 
    "DROUGHT"
excheatmatch <- c("EXCESSIVE HEAT", "EXTREME HEAT", "HEAT WAVE", "RECORD HEAT")
rawdata$EVTYPE[grep(paste(excheatmatch, collapse = "|"), rawdata$EVTYPE,
    ignore.case = TRUE)] <- "EXCESSIVE HEAT"
# "Unseasonably warm" included under Heat per 7.47 of NWSI 10-1605
heatmatch <- c("UNSEASONABLY WARM", "UNSEASONABLY HOT", "HEAT", "VERY WARM",
               "HOT", "RECORD WARM")
rawdata$EVTYPE[grep(paste(heatmatch, collapse = "|"), rawdata$EVTYPE, 
    ignore.case = TRUE)] <- "HEAT"
extcoldmatch <- c("RECORD COLD", "RECORD  COLD", "WIND CHILL")
rawdata$EVTYPE[grep(paste(extcoldmatch, collapse = "|"), rawdata$EVTYPE,
    ignore.case = TRUE)] <- "EXTREME COLD/WIND CHILL"
# "Urban flooding" included under Flood per 7.14.4 of NWSI 10-1605
floodmatch <- c("URBAN", "STREAM")
rawdata$EVTYPE[grep(paste(floodmatch, collapse = "|"), rawdata$EVTYPE,
    ignore.case = TRUE)] <- "FLOOD"
coastalmatch <- c("COASTAL FLOOD", "COASTALFLOOD")
rawdata$EVTYPE[grep(paste(coastalmatch, collapse = "|"), rawdata$EVTYPE,
    ignore.case = TRUE)] <- "COASTAL FLOOD"
marineswmatch <- c("MARINE MISHAP")
rawdata$EVTYPE[grep(paste(marineswmatch, collapse = "|"), rawdata$EVTYPE,
    ignore.case = TRUE)] <- "MARINE STRONG WIND"
tornadomatch <- c("TORNADO", "TORNDAO")
rawdata$EVTYPE[grep(paste(tornadomatch, collapse = "|"), rawdata$EVTYPE,
    ignore.case = TRUE)] <- "TORNADO"
tidematch <- c("STORM SURGE", "^TIDAL")
rawdata$EVTYPE[grep(paste(tidematch, collapse = "|"), rawdata$EVTYPE,
    ignore.case = TRUE)] <- "STORM SURGE/TIDE"

rawdata$EVTYPE <- droplevels(rawdata$EVTYPE)
```

After cleaning, the unique values of event type looks as follows:
```{r}
levels(rawdata$EVTYPE)
```


## Results
```{r}
# Split the total fatalities data by the event type.
fatalities <- aggregate(rawdata$FATALITIES, list(rawdata$EVTYPE), FUN = sum)
colnames(fatalities) <- c("EVTYPE", "FATALITIES")
# Extract data where fatalities are greater than zero.
nonzerofatals <- fatalities[fatalities$FATALITIES != 0,]
# Create a barplot of the extracted fatalities data.
plotfatal <- barplot(nonzerofatals$FATALITIES, main = "Weather Related 
                     Fatalities in US", xlab = "Event Type", ylab = "Total 
                     Fatalities", names.arg = nonzerofatals$EVTYPE, 
                     cex.names = 0.5)
```
  
The plot above shows that the weather event with the most fatalities are 
tornadoes.  
```{r}
# Split the total injuries data by the event type.
injuries <- aggregate(rawdata$INJURIES, list(rawdata$EVTYPE), FUN = sum)
colnames(injuries) <- c("EVTYPE", "INJURIES")
# Extract data where fatalities are greater than zero.
nonzeroinjurs <- injuries[injuries$INJURIES != 0,]
# Create a barplot of the extracted fatalities data.
plotinjur <- barplot(nonzeroinjurs$INJURIES, main = "Weather Related 
                     Injuries in US", xlab = "Event Type", ylab = "Total 
                     Injuries", names.arg = nonzeroinjurs$EVTYPE, 
                     cex.names = 0.5)

```
  
The plot above shows that the weather event with the most injuries are 
floods.  

```{r}
# Copy property damage to a new total damage column
rawdata$TOTALDMG <- rawdata$PROPDMG

# Apply multipliers for the property damage
rawdata$TOTALDMG[rawdata$PROPDMGEXP == "H"] <- rawdata$TOTALDMG[rawdata$PROPDMGEXP == "H"] * 100
rawdata$TOTALDMG[rawdata$PROPDMGEXP == "K"] <- rawdata$TOTALDMG[rawdata$PROPDMGEXP == "K"] * 1000
rawdata$TOTALDMG[rawdata$PROPDMGEXP == "M"] <- rawdata$TOTALDMG[rawdata$PROPDMGEXP == "M"] * 1000000

# Add crop damage with multipliers to the total damage
rawdata$TOTALDMG[rawdata$CROPDMGEXP == "H"] <- rawdata$TOTALDMG[rawdata$CROPDMGEXP == "H"] + 
    rawdata$CROPDMG[rawdata$CROPDMGEXP == "H"] * 100
rawdata$TOTALDMG[rawdata$CROPDMGEXP == "K"] <- rawdata$TOTALDMG[rawdata$CROPDMGEXP == "K"] + 
    rawdata$CROPDMG[rawdata$CROPDMGEXP == "K"] * 1000
rawdata$TOTALDMG[rawdata$CROPDMGEXP == "M"] <- rawdata$TOTALDMG[rawdata$CROPDMGEXP == "M"] + 
    rawdata$CROPDMG[rawdata$CROPDMGEXP == "M"] * 1000000

# Split the total damage cost data by the event type.
totaldamage <- aggregate(rawdata$TOTALDMG, list(rawdata$EVTYPE), FUN = sum)
colnames(totaldamage) <- c("EVTYPE", "TOTALDMG")
# Extract data where damage cost is greater than zero.
nonzerodamag <- totaldamage[totaldamage$TOTALDMG != 0,]
# Create a barplot of the extracted damage cost data.
plotdamag <- barplot(nonzerodamag$TOTALDMG, main = "Weather Related 
                     Damage in US", xlab = "Event Type", ylab = "Total 
                     Cost of Damage", names.arg = nonzerodamag$EVTYPE, 
                     cex.names = 0.5)
```
  
The plot above shows that the weather event with the greatest total damage cost
is tornadoes.